Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-07 brms_meeg.tex
updating the simulation code (now faster and more parallelised)

Understanding the temporal dynamics of cognitive processes requires
methods that can capture fast-changing neural activity with high
temporal resolution. Magnetoencephalography (MEG) and
electroencephalography (EEG) are two such methods, widely used in
cognitive neuroscience for their ability to track brain activity at the
millisecond scale. These techniques provide rich time series data that
reflect how neural responses unfold in response to stimuli or tasks. A
central goal in many M/EEG studies is to determine whether---and
when---neural responses differ across experimental conditions or
participant groups.

Here are some useful references to be discussed
(\citeproc{ref-combrisson_exceeding_2015}{Combrisson \& Jerbi, 2015};
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-frossard2021}{Frossard \& Renaud, 2021},
\citeproc{ref-frossard2022}{2022}; \citeproc{ref-gramfort2013}{Gramfort,
2013}; \citeproc{ref-hayasaka_validating_2003}{Hayasaka, 2003};
\citeproc{ref-luck_how_2017}{Luck \& Gaspelin, 2017};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al., 2019};
\citeproc{ref-pernet2015}{Pernet et al., 2015})\ldots{} See also Maris
(\citeproc{ref-maris2011}{2011}) and Rosenblatt et al.
(\citeproc{ref-rosenblatt2018}{2018}) (history of cluster-based
approaches)\ldots{} Clusters failures (\citeproc{ref-eklund2016}{Eklund
et al., 2016})\ldots{}

@@

\subsection{Previous work}\label{previous-work}

@@

where \(y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)\) denotes that
the observations \(y_{i}\) are distributed as some member of the
exponential family of distributions (e.g., Gaussian, Gamma, Beta,
Poisson) with mean \(\mu_{i}\) and scale parameter \(\phi\);
\(g(\cdot)\) is the link function, \(\mathbf{A}_{i}\) is the \(i\)th row
of a known parametric model matrix, \(\gamma\) is a vector of parameters
for the parametric terms (to be estimated), \(f_{j}\) is a smooth
function of covariate \(x_{j}\) (to be estimated as well). The smooth
functions \(f_{j}\) are represented in the model via penalised splines
basis expansions of the covariates, that are a weighted sum of basis
functions:

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). Splines
coefficients are penalised (usually through the squared of the smooth
functions' second derivative) in a way that can be interpreted, in
Bayesian terms, as a prior on the ``wiggliness'' of the function. In
other words, more complex (wiggly) basis function are penalised. To
clarify the terminology at this stage: \emph{splines} are functions
composed of simpler functions. These simpler functions are basis
functions (e.g., a cubic polynomial) and the set of basis functions is a
\emph{basis}. Each basis function gets its coefficient and the resultant
spline is the sum of these weighted basis functions.

@@

@@

Focusing on identifying onset and offset of effects (as assessed by ERP
differences or decoding performance)\ldots{} Assessing the performance
of a model-based approach (i.e., Bayesian GAMMs) to conventional methods
(multiplicity corrections or cluster-based permutation)\ldots{}

@@

Following the approach of Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see details and code in
\citeproc{ref-yeung2004}{Yeung et al., 2004}). As in Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), the signal was generated
from truncated Gaussian with an objective onset at 160 ms, a peak at 250
ms, and an offset at 342 ms. We simulated this signal for 250 timesteps
between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such
data for a group of 20 participants with 50 trials per participant and
condition (Figure~\ref{fig-eeg}).

@@

\caption{\label{fig-eeg}Averaged (mean) simulated EEG activity in two
conditions with 50 trials each, for a group of 20 participants. The
error band represents the mean plus/minus 1 standard error of the mean.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-eeg-1.pdf}

We computed the average of the ERP difference
(Figure~\ref{fig-erp})\ldots{}

\caption{\label{fig-erp}Group-level average difference between
conditions (mean +/- standard error of the mean). The `true' onset and
offset are indicated by the vertical dashed lines.}

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-erp-1.pdf}

@@

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms} (i.e., weakly informative priors). We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\). The \texttt{brms} package uses the same syntax as the
\texttt{R} package \texttt{mgcv} v 1.9-1 (\citeproc{ref-mgcv}{Wood,
2017b}) for specifying smooth effects. Figure~\ref{fig-plot-post-slope}
shows the predictions of this model together with the raw data.

@@

\subsection{Posterior probability of difference above
0}\label{posterior-probability-of-difference-above-0}

We then plot the posterior predictions together with the posterior
estimate of the slope for \texttt{condition} at each timestep
(Figure~\ref{fig-plot-post-slope}).

@@

\caption{\label{fig-plot-post-slope}Posterior estimate of the ERP in
each condition (left) or directly for the difference of ERPs (right)
according to the GAM.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-plot-post-slope-1.pdf}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0+\epsilon\)
(Figure~\ref{fig-post-prob-ratio}, left), with \(\epsilon := 0.05\),
which can be interpreted as the smallest effect size of interest. We can
also express this as the ratio of posterior probabilities (i.e.,
\(p/(1-p)\)) and visualise the timecourse of this ratio superimposed
with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}, right).

@@

\caption{\label{fig-post-prob-ratio}Left: Posterior probability of the
ERP difference (slope) being above 0 according to the GAM. Right: Ratio
of posterior probability according to the GAM (on a log10 scale).
Timesteps above threshold (10) are highlighted in green.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-post-prob-ratio-1.pdf}

@@

\subsection{Multilevel modelling using ERP summary
statistics}\label{multilevel-modelling-using-erp-summary-statistics}

Next we fit a multilevel version of the GAM. Although it is possible to
fit a GAMM at the single-trial level, we present a resource-efficient
version of the model that is fitted directly on the by-participant
summary statistics (mean and SD, similar to what is done in
meta-analysis\ldots{}

@@

We then computed the difference between the true and estimated
onset/offset of the ERP difference
(\(\text{error}:=|\hat{\theta}-\theta|\)), according to various
\texttt{k} (GAM basis dimension) and \texttt{threshold} values. Remember
that the signal is generated from a truncated Gaussian with an objective
onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms.
Figure~\ref{fig-onset-error} shows that the hierarchical GAM can
\emph{exactly} recover the true onset and offset values, given some
reasonable choice of \texttt{k} and \texttt{threshold} values (e.g., a
threshold of 20). We provide more detailed recommendations on how to set
\texttt{k} in Section~\ref{sec-basis}.

@@

\caption{\label{fig-onset-error}Error function of onset (left) and
offset (right) estimation according to various basis dimension and
threshold values (according to the multilevel GAM).}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-onset-error-1.pdf}

@@

To provide a more exhaustive assessment of the performance of the
proposed approach, we compared the ability of the GAMM to correctly
estimate the onset and offset of the ERP difference to widely-used
methods. First, we conducted mass-univariate t-tests (thus treating each
timestep independently) and identified the onset and offset of the ERP
difference as the first and last values crossing an arbitrary
significance threshold (\(\alpha = 0.05\)). We then followed the same
approach but after applying different forms of multiplicity correction
to the p-values. We compared two methods that control the false
discovery rate (FDR) (i.e., BH95, \citeproc{ref-benjamini1995}{Benjamini
\& Hochberg, 1995}; and BY01, \citeproc{ref-benjamini2001}{Benjamini \&
Yekutieli, 2001}), one method that controls the family-wise error rate
(FWER) (i.e., Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm,
1979}), and two cluster-based permutation methods (permutation with a
single cluster-forming threshold and threshold-free cluster enhancement,
TFCE, \citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The BH95,
BY01, and Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.42.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.3
(\citeproc{ref-changepoint}{Killick et al., 2022a}), and applied
directly to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}).
Figure~\ref{fig-corrections} illustrates the onsets and offsets
estimated by each method on a single simulated dataset and shows that
all methods over-estimate the true onset and under-estimate the true
offset.

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-corrections-1.pdf}

@@

The various methods were assessed using the median error and variance of
onset/offset estimates computed on 10.000 simulated datasets\ldots{}

@@

Next, we assessed the performance of the proposed approach on actual MEG
data (decoding results from \citeproc{ref-nalborczyk:inprep}{Nalborczyk
et al., in preparation}). We conducted time-resolved multivariate
pattern analysis (MVPA), also known as decoding\ldots{} As a result, we
have a timecourse of decoding performance (ROC AUC), bounded between 0
and 1, for each participant (for a total of 32 participants). Now, we
want to \emph{test} whether the group-level average decoding accuracy is
above chance (i.e., 0.5) at each timestep
(Figure~\ref{fig-decoding-data}). We fitted a similar GAM as discussed
previously, but we replaced the \(\mathrm{Normal}\) likelihood function
by a \(\mathrm{Beta}\) one to account for the bounded nature of AUC
values (between 0 and 1) (for a tutorial on Beta regression, see
\citeproc{ref-coretta2025}{Coretta \& Bürkner, 2025}). Note that
although we chose a basis dimension of \(k=50\), which seems appropriate
for the present data, this choice should be adapted according to the
properties of the modelled data (e.g., signal-to-noise ratio, prior
low-pass filtering, sampling rate, etc) and should be assessed by the
usual model checking tools (e.g., posterior predictive checks, see also
Section~\ref{sec-basis}). To distinguish signal from noise, we also
defined a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}), defined as the
chance level plus the standard deviation of the (group-level average)
decoding performance during the baseline period.

@@

We assessed the reliability of the proposed approach using a form of
permutation-based split-half reliability (as for instance in
\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}), which consisted
of the following steps. First, we created 1000 split halves of the data
(i.e., with half the participants in the original data, that is, 16
participants). For each split, we estimated the onset/offset using all
methods described previously. Third, we summarised the distribution of
onset/offset estimates using the median ``error'' (i.e., difference
between the split estimate and the estimate obtained using the full
dataset) and the variance across splits.

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-decoding-data-1.pdf}

@@

\subsection{Simulation results (bias and
variance)}\label{simulation-results-bias-and-variance}

Figure~\ref{fig-simulation-mae-variance} shows a summary of the
simulation results, revealing that the proposed approach (\texttt{BGAM})
has the lowest median error and variance for both the onset and offset
estimates.

@@

\caption{\label{fig-simulation-mae-variance}Median error and variance of
onset and offset estimates for each method. Variance is plotted on a
log10 scale for visual purposes.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-simulation-mae-variance-1.pdf}

@@

Figure~\ref{fig-onset-offset} shows the group-level average decoding
performance through time with onset and offset estimates for each
method. Overall, this figure shows that both the \texttt{Raw\ p-value}
and \texttt{FDR\ BH95} methods were extremely lenient, considering that
the decoding performance was above chance before the onset of the
stimulus (false positive) and until the end of the trial. The
\texttt{Change\ point} and \texttt{Cluster\ mass} methods were the most
conservative methods, identifying a time window from approximately +60ms
to +500ms. The \texttt{Holm}, \texttt{TFCE}, and \texttt{BGAM} methods
produced similar estimates of onset and offset, ranging from
approximately +60ms to +650ms.\footnote{It should be noted that although
  each method can produce several ``clusters'' of ``significant''
  timesteps, we only considered the first (onset) and last (offset)
  timesteps identified by each method to compute the error (difference).}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-onset-offset-1.pdf}

@@

Figure~\ref{fig-reliability} shows the median difference between the
onset and offset estimates from each data split and the onset and offset
estimates from the full dataset (x-axis) along with the variance of its
onset and offset estimates across data splits (error bar). This figure
reveals that the \texttt{BGAM} \emph{onset and offset} estimates on each
split are the closest to the estimates from the full dataset on average
(0ms difference for the onset estimate and 5ms difference for the offset
estimate). The \texttt{Raw\ p-value} method has similar performance, but
given the aberrant estimates it produces (cf.
Figure~\ref{fig-onset-offset}), the fact that it is consistent between
data splits and the full dataset is not convincing on its own. The
\texttt{Change\ point} method also has a very good performance (i.e.,
very low difference between split estimates and full estimates), but
produces too short cluster of significant decoding performance (cf.
Figure~\ref{fig-onset-offset}). Overall, the figure reveals that for all
other methods, split datasets produce later onset estimates and earlier
offset estimates (as compared to the estimates from the model fitted on
the full dataset). These results highlight that the results of any
method should be assessed using both i) good asymptotic properties on
simulated data, ii) sensible identified clusters in actual data, and
iii) reliable/stable estimates on actual data.

@@

\caption{\label{fig-reliability}Median absolute error and variance of
onset (left) and offset (right) estimates for each method.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-reliability-1.pdf}

@@

\subsection{Increasing potential
usage}\label{increasing-potential-usage}

Prepare a wrapper \texttt{R} package and show how to call it in
\texttt{Python} and integrate it with \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}) pipelines\ldots{}

@@

\subsection{Conclusions}\label{conclusions}

\ldots{}

@@

The simulation results as well as the \texttt{R} code to reproduce the
simulations are available on GitHub:
\url{https://github.com/lnalborczyk/brms_meeg}.

@@

@@

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-tidybayes}
Kay, M. (2023). \emph{{tidybayes}: Tidy data and geoms for {Bayesian}
models}. \url{https://doi.org/10.5281/zenodo.1308151}

@@

>>>>

Understanding the temporal dynamics of cognitive processes requires
methods that can capture fast-changing neural activity with high
temporal resolution. Magnetoencephalography (MEG) and
electroencephalography (EEG) are two such methods, widely used in
cognitive neuroscience for their ability to track brain activity at the
millisecond scale. These techniques provide rich time series data that
reflect how neural responses unfold in response to stimuli or tasks. A
central goal in many M/EEG studies is to determine whether and when
neural responses differ across experimental conditions or participant
groups.

Here are some useful references to be discussed
(\citeproc{ref-combrisson_exceeding_2015}{Combrisson \& Jerbi, 2015};
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-frossard2021}{Frossard \& Renaud, 2021},
\citeproc{ref-frossard2022}{2022}; \citeproc{ref-gramfort2013}{Gramfort,
2013}; \citeproc{ref-hayasaka_validating_2003}{Hayasaka, 2003};
\citeproc{ref-luck_how_2017}{Luck \& Gaspelin, 2017};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pernet2015}{Pernet et al., 2015})\ldots{} See also Maris
(\citeproc{ref-maris2011}{2011}) and Rosenblatt et al.
(\citeproc{ref-rosenblatt2018}{2018}) (history of cluster-based
approaches)\ldots{} Clusters failures (\citeproc{ref-eklund2016}{Eklund
et al., 2016})\ldots{}

@@

\subsection{Related work}\label{related-work}

@@

where \(y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)\) denotes that
the observations \(y_{i}\) are distributed as some member of the
exponential family of distributions (e.g., Gaussian, Gamma, Beta,
Poisson) with mean \(\mu_{i}\) and scale parameter \(\phi\);
\(g(\cdot)\) is the link function, \(\mathbf{A}_{i}\) is the \(i\)th row
of a known parametric model matrix, \(\gamma\) is a vector of parameters
for the parametric terms (to be estimated), \(f_{j}\) is a smooth
function of covariate \(x_{j}\) (to be estimated as well). The smooth
functions \(f_{j}\) are represented in the model via penalised splines
basis expansions of the covariates, that are a weighted sum of \(K\)
simpler, basis functions:

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). To clarify the
terminology at this stage: \emph{splines} are functions composed of
simpler functions. These simpler functions are basis functions (e.g., a
cubic polynomial) and the set of basis functions is a \emph{basis}. Each
basis function gets its coefficient and the resultant spline is the sum
of these weighted basis functions. Splines coefficients are penalised
(usually through the squared of the smooth functions' second derivative)
in a way that can be interpreted, in Bayesian terms, as a prior on the
``wiggliness'' of the function. In other words, more complex (wiggly)
basis function are penalised.

@@

Introduction to multilevel GAMs
(\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al.,
2019})\ldots{}

@@

Given the previously reported limitations of conventional methods to
precisely identify the onset and offset of M/EEG effects (e.g., ERPs,
decoding performance), we aimed to provide a model-based approach for
estimating the onset and offset of such effects. To achieve this, we
leveraged Bayesian generalised additive multilevel models (BGAMMs)
fitted via the \texttt{brms\ R} package and compared the performance of
this approach to conventional methods on both simulated and actual M/EEG
data.

@@

Following the approach of Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see code in the online supplementary materials and
details in \citeproc{ref-yeung2004}{Yeung et al., 2004}). As in
Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), the signal was
generated from truncated Gaussian with an objective onset at 160 ms, a
peak at 250 ms, and an offset at 342 ms. We simulated this signal for
250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We
simulated data for a group of 20 participants with 50 trials per
participant and condition (Figure~\ref{fig-eeg}).

@@

\caption{\label{fig-eeg}Mean simulated EEG activity in two conditions
with 50 trials each, for a group of 20 participants. The error band
represents the mean +/- 1 standard error of the mean.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-eeg-1.png}

@@

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms} (i.e., weakly informative priors). We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\) (\citeproc{ref-gabry2019}{Gabry et al., 2019};
\citeproc{ref-gelman2020}{Gelman et al., 2020}). The \texttt{brms}
package uses the same syntax as the \texttt{R} package \texttt{mgcv} v
1.9-1 (\citeproc{ref-mgcv}{Wood, 2017b}) for specifying smooth effects.
Figure~\ref{fig-plot-post-slope} shows the predictions of this model
together with the raw data.

@@

We depict the posterior predictions together with the posterior estimate
of the slope for \texttt{condition} at each timestep
(Figure~\ref{fig-plot-post-slope}).

@@

\caption{\label{fig-plot-post-slope}Posterior estimate of the EEG
activity in each condition (left) and posterior estimate of the
difference in EEG activity (right) according to the GAM.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-plot-post-slope-1.png}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0\) (Figure~\ref{fig-post-prob-ratio},
left). We can also express this as the ratio of posterior probabilities
(i.e., \(p/(1-p)\)) and visualise the timecourse of this ratio
superimposed with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}, right). A ratio of 10 means that the
probability of the difference being above 0 is 10 times higher than the
probability of the difference not being above 0, given the data, the
priors, and other model's assumptions.

@@

\caption{\label{fig-post-prob-ratio}Left: Posterior probability of the
EEG difference (slope) being above 0 according to the GAM. Right: Ratio
of posterior probability according to the GAM (on a log10 scale).
Timesteps above threshold (10) are highlighted in green.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-post-prob-ratio-1.png}

@@

\subsection{Multilevel modelling using summary
statistics}\label{multilevel-modelling-using-summary-statistics}

The previous model only included constant (fixed) effects, thus not
properly accounting for between-participant variability. We next fit a
multilevel version of the GAM. Although it is possible to fit a GAMM at
the single-trial level, we present a resource-efficient version of the
model that is fitted directly on the by-participant summary statistics
(mean and SD), similar to what is done in meta-analysis.

@@

We then compute the difference between the true and estimated
onset/offset of the EEG difference according to various \texttt{k} (GAM
basis dimension) and \texttt{threshold} values. Remember that the EEG
signal was generated from a truncated Gaussian with an objective onset
at 160 ms, a maximum at 250 ms, and an offset at 342 ms.
Figure~\ref{fig-onset-error} shows that the multilevel GAM can
\emph{almost exactly} recover the true onset and offset values, given
some reasonable choice of \texttt{k} and \texttt{threshold} values
(e.g., a threshold of 20). We provide more detailed recommendations on
how to set \texttt{k} in Section~\ref{sec-basis}.

@@

\caption{\label{fig-onset-error}Average estimation error for the onset
(left) and offset (right) according to various basis dimension and
threshold values (according to the multilevel GAM).}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-onset-error-1.png}

@@

We then compared the ability of the GAMM to correctly estimate the onset
and offset of the ERP difference to other widely-used methods. First, we
conducted mass-univariate t-tests (thus treating each timestep
independently) and identified the onset and offset of the ERP difference
as the first and last values crossing an arbitrary significance
threshold (\(\alpha = 0.05\)). We then followed the same approach but
after applying different forms of multiplicity correction to the
p-values. We compared two methods that control the false discovery rate
(FDR) (i.e., BH95, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg,
1995}; and BY01, \citeproc{ref-benjamini2001}{Benjamini \& Yekutieli,
2001}), one method that controls the family-wise error rate (FWER)
(i.e., Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}),
and two cluster-based permutation methods (permutation with a single
cluster-forming threshold and threshold-free cluster enhancement, TFCE,
\citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The BH95, BY01,
and Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.42.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.3
(\citeproc{ref-changepoint}{Killick et al., 2022}), and applied directly
to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}).
Figure~\ref{fig-corrections} illustrates the onsets and offsets
estimated by each method on a single simulated dataset and shows that
all methods over-estimate the true onset and under-estimate the true
offset.

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-corrections-1.png}

@@

The various methods were compared using the median error (i.e.,
true-estimated) and variance of onset/offset estimates computed on
10.000 simulated datasets.

@@

To provide a more exhaustive assessment of these methods, we also
evaluated the performance of the proposed approach on actual MEG data
(decoding results from \citeproc{ref-nalborczyk:inprep}{Nalborczyk et
al., in preparation}). In this study, we conducted time-resolved
multivariate pattern analysis (MVPA, also known as decoding) of MEG data
during reading tasks. As a result, we obtain a timecourse of decoding
performance (ROC AUC), bounded between 0 and 1, for each participant
(for a total of 32 participants). Next, we want to \emph{test} whether
the group-level average decoding accuracy is above chance (i.e., 0.5) at
each timestep (Figure~\ref{fig-decoding-data}). To achieve this, we
fitted a similar GAM as discussed previously, but we replaced the
\(\mathrm{Normal}\) likelihood function by a \(\mathrm{Beta}\) one to
account for the bounded nature of AUC values (between 0 and 1) (for a
tutorial on Beta regression, see \citeproc{ref-coretta2025}{Coretta \&
Bürkner, 2025}).

Note that although we chose a basis dimension of \(k=50\), which seems
appropriate for the present data, this choice should be adapted
according to the properties of the modelled data (e.g., signal-to-noise
ratio, prior low-pass filtering, sampling rate, etc) and should be
assessed by the usual model checking tools (e.g., posterior predictive
checks, see also Section~\ref{sec-basis}). To better distinguish signal
from noise, we also defined a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}), defined as the
chance level plus the standard deviation of the (group-level average)
decoding performance during the baseline period.

@@

We assessed the reliability of the proposed approach using a form of
permutation-based split-half reliability (as for instance in
\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}), which consisted
of the following steps. First, we created 1000 split halves of the data
(i.e., with half the participants in the original data, that is, 16
participants). For each split, we estimated the onset/offset using all
methods described previously. Third, we summarised the distribution of
onset/offset estimates using the median ``error'' (i.e., difference
between the split estimate and the estimate obtained using the full
dataset) and the variance across splits. This approach allows assessing
how similar the estimate of each half split is to the full dataset (thus
acting as a proxy for the population) and how variable the estimates are
(across split halves).

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-decoding-data-1.png}

@@

\subsection{Simulation study (bias and
variance)}\label{simulation-study-bias-and-variance}

Figure~\ref{fig-simulation-results} shows a summary of the simulation
results, revealing that the proposed approach (\texttt{BGAM}) has the
lowest median error and variance for both the onset and offset
estimates. The \texttt{Cluster-mass}, \texttt{Holm}, and
\texttt{Change\ point} also have good performance, but surprisingly, the
\texttt{TFCE} method has relatively bad performance for estimating the
effect offset. Unsurprisingly, the \texttt{Raw\ p-value},
\texttt{FDR\ BH95}, and \texttt{FDR\ BY01} methods show the worst
performance.

@@

\caption{\label{fig-simulation-results}Median error and variance of
onset and offset estimates for each method. The vertical dashed line
represents an unbiased estimate of the true onset/ofsset. Variance is
plotted on a log10 scale for visual purposes.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-simulation-results-1.png}

@@

Figure~\ref{fig-onset-offset} shows the group-level average decoding
performance through time with onset and offset estimates for each
method. Overall, this figure shows that both the \texttt{Raw\ p-value}
and \texttt{FDR\ BH95} methods are extremely lenient, considering that
the decoding performance is above chance before the onset of the
stimulus (false positive) and until the end of the trial. The
\texttt{Change\ point} and \texttt{Cluster\ mass} methods seem the most
conservative methods, identifying a time window from approximately +60ms
to +500ms. The \texttt{Holm}, \texttt{TFCE}, and \texttt{BGAM} methods
produce similar estimates of onset and offset, ranging from
approximately +60ms to +650ms, although the \texttt{BGAM} method seems
to result in fewer clusters.\footnote{It should be noted that although
  each method can produce several ``clusters'' of ``significant''
  timesteps, we only considered the first (onset) and last (offset)
  timesteps identified by each method to compute the estimation error.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-onset-offset-1.png}

@@

Figure~\ref{fig-reliability} shows the median difference between the
onset and offset estimates from each data split and the onset and offset
estimates from the full dataset (x-axis) along with the variance of its
onset and offset estimates across data splits (error bar). This figure
reveals that the \texttt{BGAM} \emph{onset and offset} estimates on each
split are the closest to the estimates from the full dataset on average
(0ms difference for the onset estimate and 5ms difference for the offset
estimate). The \texttt{Raw\ p-value} method has similar performance, but
given the aberrant estimates it produces (cf.
Figure~\ref{fig-onset-offset}), the fact that it is consistent between
data splits and the full dataset is not convincing on its own. The
\texttt{Change\ point} method also has a very good performance (i.e.,
very low difference between split estimates and full estimates), but
produces too short cluster of significant decoding performance (cf.
Figure~\ref{fig-onset-offset}).\footnote{As in Rousselet
  (\citeproc{ref-rousselet_using_2025}{2025}), we fixed the number of
  expected change points to two in the binary segmentation algorithm,
  thus producing always one cluster.} Overall, the figure reveals that
for all other methods, split datasets produce later onset estimates and
earlier offset estimates (as compared to the estimates from the model
fitted on the full dataset). These results highlight that the results of
any method should be assessed using both i) good asymptotic properties
on simulated data, ii) sensible identified clusters in actual data, and
iii) reliable/stable estimates on actual data.

@@

\caption{\label{fig-reliability}Median error and median absolute
deviation of the error for the onset (left) and offset (right) estimates
according to each method. Methods are ordered from lowest (top) to
highest (bottom) median absolute error (separately for the onset and
offset estimates).}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-reliability-1.png}

@@

@@

@@

The simulation results as well as the \texttt{R} code to reproduce the
simulations are available on GitHub:
\url{https://github.com/lnalborczyk/brms_meeg}. The \texttt{neurogam\ R}
package is available at \url{https://github.com/lnalborczyk/neurogam}.

@@

\bibitem[\citeproctext]{ref-RJ-2021-048}
Bengtsson, H. (2021). A unifying framework for parallel and distributed
processing in r using futures. \emph{The R Journal}, \emph{13}(2),
208--227. \url{https://doi.org/10.32614/RJ-2021-048}

@@

\bibitem[\citeproctext]{ref-doParallel}
Corporation, M., \& Weston, S. (2022). \emph{{doParallel}: Foreach
parallel adaptor for the {``{parallel}''} package}.
\url{https://CRAN.R-project.org/package=doParallel}

@@

\bibitem[\citeproctext]{ref-gabry2019}
Gabry, J., Simpson, D., Vehtari, A., Betancourt, M., \& Gelman, A.
(2019). Visualization in Bayesian work{fl}ow. \emph{Journal of the Royal
Statistical Society: Series A (Statistics in Society)}, \emph{182}(2),
389--402. \url{https://doi.org/10.1111/rssa.12378}

@@

\bibitem[\citeproctext]{ref-gelman2020}
Gelman, A., Vehtari, A., Simpson, D., Margossian, C. C., Carpenter, B.,
Yao, Y., Kennedy, L., Gabry, J., Bürkner, P.-C., \& Modrák, M. (2020).
Bayesian workflow. \emph{arXiv:2011.01808 {[}Stat{]}}.
\url{http://arxiv.org/abs/2011.01808}

@@

\bibitem[\citeproctext]{ref-glue}
Hester, J., \& Bryan, J. (2024). \emph{{glue}: Interpreted string
literals}. \url{https://CRAN.R-project.org/package=glue}

@@

\bibitem[\citeproctext]{ref-tidybayes}
Kay, M. (2024). \emph{{tidybayes}: Tidy data and geoms for {Bayesian}
models}. \url{https://doi.org/10.5281/zenodo.1308151}

@@

\bibitem[\citeproctext]{ref-easystats}
Lüdecke, D., Ben-Shachar, M. S., Patil, I., Wiernik, B. M., Bacher, E.,
Thériault, R., \& Makowski, D. (2022). {easystats}: Framework for easy
statistical modeling, visualization, and reporting. \emph{CRAN}.
\url{https://doi.org/10.32614/CRAN.package.easystats}
